Formalisation of Session-based Lambda Calculus Type Systems

### 2014-11-05 proposal/proposal.tex
Added beginnings of typing rules for GV; need opening and substitution.
Specifically, need opening for the abstraction case (and the other
binding rules). These definitions need to respect linearity.
Added introductory paragraphs and some references.

\usepackage{graphicx}
\usepackage[style=alphabetic,natbib=true]{biblatex}
\addbibresource{proposal}

@@

computing distributed communication rise of internet networked communication
the protocols have an effect on the communication and if the protocol is wrong
distributed communicatin now a propenent is most peoples' daily lives
whether it be posting pictures on social media websites sharing videos
or even so much as sending emails all require a well-defined protocol to
specify the type and order of messages to be sent and received between the two
or more communicating partners
mainstream programming languages used to build such systems often fail to meet
the strong guarantees that the correct type of messages will be sent or that
the ordering between messages is preserved
the rise of e-commerce and increasing reliance on software systems in the
areas of aviation/transport mean that much confidence in these systems is placed thus it is of great importance that there exist strong guarantees of correctness

@@

The aim of this project is to develop a framework for formalising lambda
calculus type systems involving session types using interactive theorem
proving (in particular, the system Coq). The motivation for the framework is
that it will enable a whole class of related languages and type systems to be
formalised with little extra proof effort. Inspired by previous work on a
mechanised framework of pi calculus type systems~\cite{Gay:2001:FFP}, we shift
to the lambda calculus to provide a basis on which session types for
mainstream programming languages can be studied whilst taking advantage of
previous proof effort and specialising only for those parts unique to the
particular system.

From the generic framework, a number of directions can be explored including
session polymorphism, session subtyping, and permitting different forms of
aliasing within the type system. The majority of previous session type systems
for mainstream programming languages has focussed exclusively on linear type
systems. One disadvantage of a linear type system is the need to re-bind
objects after each operation~\cite{Gay:2010:LTT}. Principally, we intend to
instantiate the framework with a type system that provides the benefits of
aliasing with the guarantees of linearity, termed ``adoption and focus''. The
``adoption and focus'' type system is non-trivial requiring special language
features, e.g. to manage capabilities which can be thought as a form of object
liveness mechanism, so the framework design must support such flexibility by,
e.g allowing the typing judgements to be extended such that the capabilities
can be included in soundness results. The instantiation will demonstrate that
the framework supports a wide range of languages and type systems providing
researchers the ability to harness the power of formal verification in the
design, implementation and documentation of session-based type systems.

@@

System F-pop, session polymorphism formalised in Coq, Simon's Isabelle/HOL
development.

@@

>>>>

\usepackage{graphicx}
\usepackage[style=alphabetic,natbib=true]{biblatex}
\addbibresource{proposal.bib}

@@

Distributed communication systems are becoming increasing prevalent in modern society. To facilitate communication between concurrently executing agents requires complex network protocols to define the type and sequencing of messages. Typically these protocols are implemented in systems programming languages, such as C, which provide no correctness guarantees of the implementation. Furthermore, programming in these languages is often extremely error-prone, as one has a myriad of unsafe features at one's disposal. As more businesses begin to conduct transactions over the Web, it is imperative that they can reason about the communication protocols upon which they rely.

Encoding communication protocols as \textit{session types} is one approach to achieve guarantees about correctness. One particular variant of session types are \textit{binary} session types~\footnote{There are also \textit{multiparty} session types, but I do not discuss them here.}. A binary session type specifies communication between two agents, and it specifies the order and type of messages to be sent or received by an agent in the system. The two agents have dual session types such that, for example, when one is expecting to receive a message the other is expecting to send a message; both session types agree on the type of message to be exchanged. In this way, correctness of communication can be statically assured by providing communicating agents with \textit{channels} that have dual session types. A channel is defined as a bi-directional conduit for exchanging messages, not unlike a socket in network programming. In contrast to sockets, however, a channel is associated with an explicit type which provides a description of its protocol. Thus, in session-based systems a channel is usually provided as a primitive type.

@@

The aim of this project is to develop a framework for formalising lambda
calculus type systems involving session types using interactive theorem
proving (in particular, the system Coq). The motivation for the framework is
that it will enable a whole class of related languages and type systems to be
formalised with little extra proof effort. Inspired by previous work on a
mechanised framework for pi calculus type systems~\cite{Gay:2001:FFP}, we
shift to the lambda calculus to provide a basis on which session types for
mainstream programming languages can be studied whilst taking advantage of
previous proof effort and specialising only for those parts unique to the
particular system.

From the generic framework, a number of directions can be explored including
session polymorphism, session subtyping, and permitting different forms of
aliasing within the type system. The majority of previous session type systems
for mainstream programming languages has focussed exclusively on linear type
systems. One disadvantage of a linear type system is the need to re-bind
objects after each operation~\cite{Gay:2010:LAST}. Principally, we intend to
instantiate the framework with a type system that provides the benefits of
aliasing with the guarantees of linearity, termed ``adoption and focus''. The
``adoption and focus'' type system is non-trivial requiring special language
features, e.g. to manage capabilities which can be thought as a form of object
liveness mechanism, so the framework design must support such flexibility by,
e.g allowing the typing judgements to be extended such that the capabilities
can be included in soundness results. The instantiation will demonstrate that
the framework supports a wide range of languages and type systems providing
researchers the ability to harness the power of formal verification in the
design, implementation and documentation of session-based type systems.

@@

System F-pop~\cite{Mazurak:2010:LLT}
session polymorphism formalised in Coq, Simon's Isabelle/HOL
development.
Philip Wadler's Propositions as Sessions~\cite{Wadler:2012}
Simon J. Gay and Vasco Vasconcelos (LAST)~\cite{Gay:2010:LAST}
Lolliproc~\cite{Mazurak:2010:LCC}
PhD thesis~\cite{Mazurak:2013:LPP}

@@

HOAS Coq formalisation using Metatheory library from UPenn.
Explain why this approach was taken versus intrinsic encoding of the terms
